Making multiple-locus association mapping on a genome-wide scale routine

### 2017-10-24 AM+draft.tex
25Oct17 Methods - Dimension reduction

%%\usepackage{graphicx}

%%\linespread{1.5}
%%\addtolength{\oddsidemargin}{-2.5cm}
%\addtolength{\evensidemargin}{-3cm}
%%\addtolength{\textwidth}{2.5cm}
\begin{document}

@@

Phenotypic and genotypic data were collected on a large study population of commercially 
available outbred mice. The aim of the study was to map, through association, the genetic basis
of complex traits in mice. The construction of the population and collection of data are described in detail in 
\cite{Nicod:2016aa}.  Briefly, the population consists of 2073 adult outbred mice from Charles River Laboratories. Phenotypic data
were obtained from 200 traits. These traits fell into the categories of behaviour, tissue, and physiology. Genotype 
data were collected from 7, 073, 398 SNPs.  A reduced set of 359, 559 SNP across the 19 
autosomes and sex chromosome was then identified, upon which all genome-wide analyses were based.  These SNP tagged all other SNP with MAF $> 0.1\%$
and LD $r^2 > 0.98$.  Missing data?

@@

The Eagle approach for multiple-locus association mapping is based on building a linear mixed model, iteratively, via forward 
selection.  Suppose $s$ iterations of our model building process have already been performed. This means $s$ snp-trait 
associations have been identified.  It also means that $s$ separate genomic regions of interest have been found.

To perform the $s+1$th  iteration of 
our model building procedure, we do the following. 
First, we fit the current model to the data. 
Let $S = \{ S_1, S_2, \ldots, S_s\}$ be a set of ordinal numbers where $S_k$ is the $S_k$th ordered snp that was 
selected in the $k$th iteration. For example, if three iterations of our model building procedure 
have been performed and say the 500023rd, 15th, and 420th, 
snp were selected, then $S=\{500023, 15, 420\}$. 
Let $y^{(n \; \times \;1)}$ be a vector containing the quantitative trait data. 
Let $M^{(n \; \times \; L)} = [m_1 m_2 \ldots m_L]$ be a matrix containing the genotype data which have been collected 
from $L$ loci that span the genome.  It is common for the columns of this matrix to be in map order but this is not a requirement. 
The vector $m_j^{(n \; \times \; 1)}$ contains the genotypes, from the $n$ individuals,  for the $j$th snp. 
The genotypes are coded as -1, 0, and 1 corresponding to snp genotypes AA, AB, and BB, respectively.

@@

The Eagle approach has its foundations in the work of XXXX and XXXXX. Here, the authors developed a more 
general approach and have used it for mapping QTL in linkage mapping studies in plants (refs).

\bibliographystyle{naturemag}
\bibliography{biblibrary}

>>>>

\begin{document}

@@

Phenotypic and genotypic data were collected on a large study population of commercially 
available outbred mice. The aim of the study was to map, through association, the genetic basis
of complex traits in mice. The construction of the population and collection of data are described in detail in 
\cite{nicod2016genome}.  Briefly, the population consists of 2073 adult outbred mice from Charles River Laboratories. Phenotypic data
were obtained from 200 traits. These traits fell into the categories of behaviour, tissue, and physiology. Genotype 
data were collected from 7, 073, 398 SNPs.  A reduced set of 359, 559 SNP across the 19 
autosomes and sex chromosome was then identified, upon which all genome-wide analyses were based.  These SNP tagged all other SNP with MAF $> 0.1\%$
and LD $r^2 > 0.98$.  Missing data?

@@

The Eagle approach for multiple-locus association mapping is based on building a linear mixed model, iteratively, via forward 
selection.  It is a simplification of a whole-genome method for mapping quantitative trait loci in classic linkage studies \cite{verbyla2007analysis}.
Suppose $s$ iterations of our model building process have already been performed. This means $s$ snp-trait 
associations have been identified.  It also means that $s$ separate genomic regions of interest have been found.  
To perform the $s+1$th  iteration of 
our model building procedure, we do the following.

First, we fit the current model to the data. 
Let $S = \{ S_1, S_2, \ldots, S_s\}$ be a set of ordinal numbers where $S_k$ is the $S_k$th ordered snp that was 
selected in the $k$th iteration. For example, if three iterations of our model building procedure 
have been performed and say the 500023rd, 15th, and 420th, 
snp were selected, then $S=\{500023, 15, 420\}$. 
Let $y^{(n \; \times \;1)}$ be a vector containing the quantitative trait data. 
Let $M^{(n \; \times \; L)} = [m_1 m_2 \ldots m_L]$ be a matrix containing the genotype data which have been collected 
from $L$ loci that span the genome.  It is common for the columns of this matrix to be in map order but this is not a requirement. 
The vector $m_j^{(n \; \times \; 1)}$ contains the genotypes, from the $n$ individuals,  for the $j$th snp. 
The genotypes are coded as -1, 0, and 1 corresponding to snp genotypes AA, AB, and BB, respectively.

@@

\subsubsection{Reducing the dimension of the model}

In practice, estimating the parameters of (\ref{eq2}) can be demanding, computationally. 
The vector $a_{-S}$ has $L-s$ random effects where in modern genome-wide association studies, 
$L$, the number of snp, can be extremely large.  An alternative model is given by 
Verbyla. They show how to reformulate (\ref{eq2}) as 
.....
where ..... and XXX is calculated via s. v. d.  Although it may not be obvious, the two models are equivalent, 
having identical variance structures. Yet, the computational cost of model (\ref{eq3}) compared to 
model \ref{eq2}, in most modern studies,  is much less because the sample size $n$ is 
often much smaller than the number of snp $L$.

Verbyla go on to show how to recover $\widetilde{a}$ from estimates from model (3) with 
\begin{equation}
\widetilde{a} = \left [ M_{-S}^T (M_{-S} M_{-S}^T)^{-1/2} \right ] \widetilde{a}^*
\end{equation}
and its variance matrix with 
\begin{equation}
\label{eq4}
\textrm{var}(\widetilde{a}) = M_{-S}^T (M_{-S} M_{-S}^T)^{-1/2} \textrm{var}(\widetilde{a}^*) (M_{-S} M_{-S}^T)^{-1/2} M_{-S}
\end{equation}
These values are needed to calculate the score statistic $t_j^2$ for identifying the snp in strongest association with the trait. 
Fortunately, when calculating $t_j^2$, only the diagonal elements of the variance matrix are needed which simplifies the  calculation. 
of (\ref{eq4}).

Fortunately

Since 
$\widetilde{a}$ and $\textrm{var} (\widetilde{a}) $ are needed  to calculate $t_j^2$, as described in 
Verbyla \cite{verbyla2012rwgaim, verbyla2014whole}, 
\begin{equation}
\widetilde{a} = \left [ M_{-S}^T (M_{-S} M_{-S}^T)^{-1/2} \right ] \widetilde{a}^*
\end{equation}
with variance matrix
\begin{equation}
\textrm{var}(\widetilde{a}) = M_{-S}^T (M_{-S} M_{-S}^T)^{-1/2} \textrm{var}(\widetilde{a}^*) (M_{-S} M_{-S}^T)^{-1/2} M_{-S}
\end{equation}
Fortunately, only the diagonal elements of the variance matrix are needed which simplifies its calculation.

It is not uncommon in genome wide association studies for the number of 
snp ($L$) to be many orders of magnitude larger than the sample size ($n$). 
As a result, estimating $a$ in (\ref{eq2}) can be demanding, computationally. 
We avoid having to estimate $a$ directly by instead reformulating (\ref{eq2}) as
the dimension reduced model 
\begin{equation}
\label{eq3}
u_g = \sum_{k=1}^s  m_{S_k} a_{S_k} + (M_{-S} M_{-S}^T)^{1/2} a^*_{-S}
\end{equation}
where $a^*$ is a $n \times 1$ vector of random effects that follows $N(0, \sigma_a^2 I^{(n \; \times \;  n)})$, and 
$(M_{-S} M_{-S}^T)^{1/2}$ can be calculated via single value decomposition (ref). 
Models (\ref{eq2})  and (\ref{eq3}) are equivalent models 
\cite{verbyla2012rwgaim, verbyla2014whole}. They have the same variance structure, 
despite the random effect in (\ref{eq3}) being only of size $n$.  This allows us to use (\ref{eq3}) in place of 
(\ref{eq2}) in the above described Eagle approach, affording us considerable computational benefits. Since 
$\widetilde{a}$ and $\textrm{var} (\widetilde{a}) $ are needed  to calculate $t_j^2$, as described in 
Verbyla \cite{verbyla2012rwgaim, verbyla2014whole}, 
\begin{equation}
\widetilde{a} = \left [ M_{-S}^T (M_{-S} M_{-S}^T)^{-1/2} \right ] \widetilde{a}^*
\end{equation}
with variance matrix
\begin{equation}
\textrm{var}(\widetilde{a}) = M_{-S}^T (M_{-S} M_{-S}^T)^{-1/2} \textrm{var}(\widetilde{a}^*) (M_{-S} M_{-S}^T)^{-1/2} M_{-S}
\end{equation}
Fortunately, only the diagonal elements of the variance matrix are needed which simplifies its calculation.

%%\bibliographystyle{naturemag}
\bibliography{biblibrary}
